Liveness-Based Garbage Collection for Lazy Languages

### 2015-05-09 paper/paper.tex
commented unwanted input
Liveness

A variable is {\em live} if  there is a possibility of its value being
used in  future computations  and dead if  it is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness  of  pointers  including  fields  of  $\CONS$  cells.   Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second and  third members of $x$.   A {\em liveness  environment} is a
mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
expressed  as a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},           y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.    In  this   notation,  $x   \mapsto  \{\epsilon\}$
represents access using $x$ itself and $x \mapsto \{\}$ indicates that
$x$ is dead.  We associate  a liveness environment with program points
denoted by $\pi$ and $\psi$.  The liveness at $\pi:e$  is the liveness just before executing
$e$.

A  notion related  to  liveness is  {\em  demand}.  The  demand on  an
expression  $e$  is again  a  set  of  access paths---that  subset  of
$\{\acar,\acdr\}^\ast$ which  the context of $e$ may  explore of $e$'s
result.   To see  the need  for demands,  consider the  let expression
$\pi:(\LET~x\leftarrow (\CDR~y)~\IN~ \pi':\RETURN~x)$. Assume that the
context of  this expression produces the liveness  $[x \mapsto \lbrace
  \epsilon, \acar  \rbrace]$ at $\pi'$.   Due to the  \LET\ definition
which  binds $x$  to $(\CDR~y)$,  the liveness  of $x$  at  $\pi'$ now
becomes  the  demand on  $(\CDR~y)$.   This,  in  turn, generates  the
liveness $\lbrace y.\epsilon,  y.\acdr, y.\acdr\acar \rbrace$ at $\pi$
(the set of $y$-rooted accesses required to explore $\lbrace \epsilon,
\acar  \rbrace$  paths  of  the  result  of  $(\CDR~y)$)\footnote{The
  classical analogue  is strong liveness  analysis, where $y$  and $z$
  have  a  liveness  of  $\lbrace\epsilon\rbrace$  at  the  entry  $n:
  x:=y+z$,    if   and    only   if    $x$   has    a    liveness   of
  $\lbrace\epsilon\rbrace$ at  exit of  $n$.  In our  terminology, the
  liveness of $x$  at exit from $n$ becomes the  demand on $y+z$. This
  generates the  liveness $\lbrace y.\epsilon,  z.\epsilon \rbrace$ at
  the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).

@@

 In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
 generates no  liveness.  Notice that this is  different from liveness
 of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
 evaluated even for the null  demand, and this would require an access
 using  $x$.   A  non-null   demand  of  $\sigma$  on  (\CDR~$x$),  is
 transformed  to  the  liveness  $\{x.\epsilon, x.\acdr\sigma\}$.   In  an
 opposite  sense, the  demand of  $\acdr\sigma$ on  (\CONS~$y$~$z$) is
 transformed to  the demand  $\sigma$ on $z$.   Note that  \CONS\ does
 not, by itself,  dereference  its arguments. The rules for (\PRIM~x~y)
 and (\NULLQ~x) are on similar lines. Constants generate no liveness.

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.   %% To deal  with this,  we introduce  an analysis  which is
%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
%%   path-sensitive means certain prooperties  holding along a path only,
%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
demand $\sigma$.  Clearly, this is also the demand on the return value
$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
in traditional liveness analysis.

\begin{figure*}[t!]
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_5\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_6\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_7\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[$ 
              $ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\; \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace ]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $[~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
                $~~\psi_1 \mapsto \lbrace ep_1 \rbrace, $ \
                $ \psi_2 \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_5 \mapsto \lbrace ep_3 \rbrace, $ \\
                $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace\rbrack$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace, $ \\
                  $~~~ep_4 \mapsto \lbrace ep_4 \rbrace,\ldots~\rbrack$  
                \end{tabular}
        \end{tabular}
      } 
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
beyond the scope of the definition of \py.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \py\ is yet to
come into existence at $\pi_6$  and its liveness will not be consulted
by the garbage collector at this point.
In an eager language, \py\ would not have been live at $\psi_3$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
$(1+~\py)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
every program  point from $\psi_3$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider  paths  from the root of  this tree to
program  points  which  trigger  a evaluation\footnote{Program points
  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em
  e-paths}. \comment{Update the e-path labels}  For the
function \length, there are three such paths $ep_1 = \pi_1, \pi_2, \psi_1$,
$ep_2 = \pi_1,  \pi_2, \psi_2$ and  $ep_3  =  \pi_1, \pi_2,  \pi_3,
\pi_4, \pi_5,  \psi_3$.

The analysis consists of two maps:  The map $\cal{M}$ maps a e-path to
a liveness environment arising out  of all uses of variables along the
path, and  a second map $\cal{P}$ maps  a program point to  the set of
e-paths sharing the point.  If a program point happens to fall on more
than one such path, then the liveness environment of the program point
is  the  variable-wise  union  of  the liveness  environments  of  the
individual paths.  In the example program, ${\cal M}(\psi_1)= \ldots$,
${\cal  M}(\psi_2)=\ldots$  and  ${\cal  M}(\psi_3)=  \ldots$.   Since
${\cal  P}(pi_4)$ is  $\lbrace  \psi_1, \psi_2,  \psi_3 \rbrace$,  the
liveness of $pi_4$ is \ldots.  In summary, our propagation of liveness
is  flow-insensitive  but  path-sensitive.   In the  sequel  we  shall
describe  how to obtain  $\cal{M}$.  The  formulation of  $\cal{P}$ is
easy and we do not elaborate it any further.

@@

In an eager language, \pw\ would not have been live after $\pi_7$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
$(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
every program  point from $\pi_8$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
$ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
 and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
  \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
  M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
is created and added to  the union.  Also notice that the consequences
of  laziness---the entire  expression including  the condition  is not
evaluated if  the demand  on it is  $\emptyset$.  This results  in the
empty map denoted by $\emptyset_{\cal{M}}$.

>>>>

A variable is {\em live} if  there is a possibility of its value being
used in  future computations  and dead if  it is definitely  not used.
Heap-allocated data needs a richer model of liveness which talks about
liveness  of  pointers  including  fields  of  $\CONS$  cells.   Using
$\acar$, $\acdr$  to represent access using $\CAR$  and $\CDR$ fields,
the  liveness  of the  structure  reachable  from  a variable  can  be
represented by a set of {\em access paths} i.e.  prefix-closed strings
from $\{\acar,\acdr\}^\ast$.   As an  example, if $x$  is a  list with
liveness      $\{\epsilon,     \acdr,      \acdr\acar,     \acdr\acdr,
\acdr\acdr\acar\}$, then future computations  can only refer up to the
second and  third members of $x$.   A {\em liveness  environment} is a
mapping from variables to subsets of $\{\acar,\acdr\}^\ast$, but often
expressed  as a set,  for example  by writing  $\{x.\epsilon, x.\acdr,
x.\acdr\acdr,        y.\epsilon\}$        instead        of        $[x
  \mapsto\{\epsilon,\acdr,\acdr\acdr\},           y\mapsto\{\epsilon\},
  z\mapsto\{\}]$.    In  this   notation,  $x   \mapsto  \{\epsilon\}$
represents access using $x$ itself and $x \mapsto \{\}$ indicates that
$x$ is dead.  We associate  liveness environments with program points
(denoted by $\pi$ and $\psi$).  The liveness at $\pi:e$  is the liveness just before executing
$e$.

A  notion  related  to  liveness  is  {\em  demand}.   While  liveness
represents the  future use of variables, demands  represent the future
use of the  values of expressions. The demand on  an expression $e$ is
again a  set of  access paths---that subset  of $\{\acar,\acdr\}^\ast$
which the context of $e$ may explore of $e$'s result.  To see the need
for  demands,  consider   the  let  expression  $\pi:(\LET~x\leftarrow
(\CDR~y)~\IN~  \pi':\RETURN~x)$.  Assume  that  the  context  of  this
expression produces  the liveness $[x \mapsto  \lbrace \epsilon, \acar
  \rbrace]$ at $\pi'$.  Due to the \LET\ definition which binds $x$ to
$(\CDR~y)$, the  liveness of $x$ at  $\pi'$ now becomes  the demand on
$(\CDR~y)$.    This,  in   turn,  generates   the   liveness  $\lbrace
y.\epsilon,  y.\acdr,  y.\acdr\acar  \rbrace$  at $\pi$  (the  set  of
$y$-rooted  accesses  required  to  explore $\lbrace  \epsilon,  \acar
\rbrace$  paths of  the result  of  $(\CDR~y)$)\footnote{The classical
  analogue  is strong  liveness analysis,  where  $y$ and  $z$ have  a
  liveness of  $\lbrace\epsilon\rbrace$ at  the entry $n:  x:=y+z$, if
  and only if  $x$ has a liveness of  $\lbrace\epsilon\rbrace$ at exit
  of $n$.   In our terminology, the  liveness of $x$ at  exit from $n$
  becomes the  demand on $y+z$.  This, in turn,  generates the  liveness $\lbrace
  y.\epsilon, z.\epsilon \rbrace$ at the entry of $n$.}.
%% Exploring $\lbrace \epsilon, \acar \rbrace$ of the
%% result of $(\CDR~y)$ first requires access using $y$ ( and the \CDR\ field of the cons cell pointed by
%% $y$.
Note that  liveness refers to  access using variables and  fields, and
not  to $\CONS$  cells (i.e.\  to edges  in the  memory graph,  not to
locations themselves).
\begin{figure*}[t!]
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{.27cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{.29cm}    $\pi_3\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{.34cm}   $\pi_4\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{.34cm} $\pi_5\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{.34cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_6\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_7\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_8\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        %%} 
    \end{boxedminipage}

    &

    \begin{boxedminipage}{.46\textwidth}

        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[~ep_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,~ep_2 \mapsto \lbrace \rbrace , $\\
              $~~ep_3 \mapsto \lbrace \pz.\sigma_{\length},
                \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, $\\
              $~~~~~~~~~~~~~~\;\;
                \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}
                \rbrace~\rbrack$ \\
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $~[
                  ~\pi_1 \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_2 \mapsto \lbrace ep_1, ep_2, ep_3
                  \rbrace, $\\
                $~~\psi_1 \mapsto \lbrace ep_1 \rbrace, $ \
                $ \psi_2 \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_5 \mapsto \lbrace ep_3 \rbrace, $ \\
                $~~\pi_6 \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_7 \mapsto \lbrace ep_3 \rbrace,~ \psi_3 \mapsto \lbrace ep_3 \rbrace~\rbrack$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}$, $\pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}$, $\ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $~\lbrack$~$\pi_6 \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_7 \mapsto \lbrace ep_4 \rbrace, $ \
                  $\pi_8 \mapsto \lbrace ep_4 \rbrace,~\psi_4 \mapsto \lbrace ep_4 \rbrace~\rbrack$  
                \end{tabular}
        \end{tabular}

    \end{boxedminipage} 
    %% \end{ tabular}}
  \end{tabular}
%\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.{\color{red} Fix the
    formatting of the table on the right}}\label{fig:mot-example2}    
\end{figure*}

@@

 In each  rule defining $\mathit{ref}$, the  null demand ($\emptyset$)
 generates no  liveness.  Notice that this is  different from liveness
 of  eager  languages,  where,   for  instance,  (\CDR~$x$)  would  be
 evaluated  even for  the null  demand. This  would require  an access
 using $x$ resulting in  $x.\epsilon$ being live.  A non-null demand of
 $\sigma$ on (\CDR~$x$), is transformed to the liveness $\{x.\epsilon,
 x.\acdr\sigma\}$.  In an opposite  sense, the demand of $\acdr\sigma$
 on  (\CONS~$y$~$z$) is  transformed to  the demand  $\sigma$  on $z$.
 Note that \CONS\ does not,  by itself, dereference its arguments. The
 rules for (\PRIM~x~y) and  (\NULLQ~x) are on similar lines. Constants
 generate no liveness.

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second reason why 
liveness analysis for lazy and eager languages are different---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.   %% To deal  with this,  we introduce  an analysis  which is
%% flow-insensitive  but path-sensitive\comment{Checked  the terminology,
%%   path-sensitive means certain prooperties  holding along a path only,
%%   http://en.wikipedia.org/wiki/Data-flow\_analysis} into our analysis.
To see  the consequences of this, consider  an analysis  of the body  of $\length$  with a
demand $\sigma$.  Clearly, this is also the demand on the return value
$z$   giving  the   liveness  at   the  program   point   $\psi_3$  as
$\{\pz.\sigma\}$.  The liveness at $\psi_3$ is propagated backwards as
in traditional liveness analysis.

@@

Since   \pz\  is   $(1  +   \py)$,  the   liveness  of   $\sigma$  for
\pz\ translates into a demand  of $\sigma$ on $(1+\py)$, which in turn
generates a  liveness of $\{\py.\epsilon\}$ at  $\pi_5$.  We now make
the following observations:
\begin{enumerate}
\item While the liveness of \py\  should be killed at $\pi_4$ since it
  is  beyond the scope  of the  definition of  \py, we  do not  do so.
  However, this  does not cause any  imprecision in the way  of the GC
  marking more cells as live, since \py\ is yet to come into existence
  at $\pi_4$,  and its liveness will  not be consulted  by the garbage
  collector at  this point. 
\item In an eager language, \py\ would not have been live at $\psi_3$,
  a program point  beyond its last use.  In  a lazy language, however,
  the  definition $\pz  \leftarrow  (1+\py)$, does  not  result in  an
  evaluation  of  $(1+\py)$;  instead   a  closure  is  created.   The
  evaluation  actually takes  place while  evaluating $(\RETURN~\pz)$.
  We therefore  record  the liveness of $\py$ at the program  
  point $\psi_3$,  indeed at every  program point from $\psi_3$  up to
  the beginning  of the program (since  we do not  kill liveness).
\end{enumerate}

We thus regard  the body of the function as a  tree and consider paths
from  the  root  of this  tree  to  program  points which  trigger  an
evaluation\footnote{Program  points which  trigger  evaluation  are
  labelled with  $\psi$ {\color {red} ---slightly late in the day}}---\SIF\ conditions and  \RETURN\ expressions.
We call such paths {\em e-paths}.  For the function \length, there are
three such paths $ep_1 = \langle \pi_1, \pi_2, \psi_1\rangle$, $ep_2 =
\langle\pi_1, \pi_2, \psi_2\rangle$ and  $ep_3 = \langle \pi_1, \pi_2,
\pi_3, \pi_4, \pi_5, \psi_3\rangle$.

Given  a function  $\mathit{f}$ and  a demand  $\sigma$,  the liveness
analysis   comes   the    map   $\cal{M}_{\mathit   f}^{\sigma}$   (or
$\cal{M}_{\mathit  f}$ when  the  demand $\sigma$  is  clear from  the
context) maps  a e-path to a  liveness environment arising  out of all
uses of variables  along the path. A second  map $\cal{P}_{\mathit f}$
maps a  program point to the set  of e-paths sharing the  point.  If a
program point  happens to fall  on more than  one such path,  then the
liveness environment  of the program point is  the variable-wise union
of the liveness environments of  the individual paths.  In the example
program,  since ${\mathcal P}_{\length}(\pi_1)  = \lbrace  ep_1, ep_2,
ep_3\rbrace$,  the liveness  at $\pi_1$  is  ${\cal M_{\length}}(ep_1)
\cup {\cal  M_{\length}}(ep_2)\cup {\cal M_{\length}}(ep_3)$.   In the
sequel   we   shall  describe   how   to   obtain  $\cal{M}$   through
$\mathcal{L}$.  The  formulation of  $\cal{P}$ is easy  and we  do not
elaborate it any further.

{\color{red}  \hrule \hrule \hrule  \medskip }

@@

%% \mycomment{In an eager language, \pw\ would not have been live after $\pi_7$, the
%% point of  its last  use. In a  lazy language, however,  the definition
%% $\pu  \leftarrow (\CDR~\pw)$,  does  not result  in  an evaluation  of
%% $(\CDR~\pw)$; instead  a closure is created.   The evaluation actually
%% takes place while evaluating  $(\RETURN~\pu)$. Thus it is necessary to
%% record the liveness  of $\pw$ at the program  point $\pi_8$, indeed at
%% every program  point from $\pi_8$ up  to the beginning  of the program
%% (since  we do  not kill  liveness).  We  thus regard  the body  of the
%% function as  a tree and consider paths  from the root of  this tree to
%% program  points  which  trigger  a evaluation---\SIF\  conditions  and
%% \RETURN\  expressions.  We  call such  paths {\em  e-paths}.   For the
%% program shown, there are three such paths $ep_1 = \lbrace\pi_1, \pi_2, \pi_4 \rbrace$, 
%% $ep_2  =  \lbrace \pi_1, \pi_2,  \pi_5,  \pi_6,  \pi_7 \rbrace$,
%%  and $ep_3 = \lbrace \pi_3, \pi_2, \pi_1 \rbrace$.

%% The analysis  consists of  two maps: The  map $\cal{M}$ maps  a return
%% path to a common liveness environment, and a second map $\cal{P}$ maps
%% a program  point to the set of  e-paths on which  the point lies.
%% If a  program  point happens to fall  on more than one  such path, then
%% the  liveness environment of  the program  point is  the variable-wise
%% union of  the liveness environments  of the individual paths.   In the
%% example program, ${\cal M}(ep_1)= \lbrace \rbrace$, ${\cal M}(ep_2)= \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace  $ and
%% ${\cal M}(ep_3)= \lbrace \px.\epsilon, \xl.\epsilon \rbrace $.
%% Since  ${\cal P}(\pi_1)$ is  $\lbrace ep_1,ep_2, ep_3 \rbrace$, 
%% the liveness  of $\pi_1$ is $\lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon,  \px.\epsilon  \rbrace$.  In summary,
%% our propagation of liveness is flow-insensitive but path-sensitive.
%% In the sequel
%% we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
%% $\cal{P}$ is easy and we do not elaborate it any further.

%% Consider  the $\mathcal{L}$-rules for  {\LET}, {\SIF},  and {\RETURN}.
%% The  expression $\pi:(\RETURN~x)$  with  a demand  $\sigma$ forces  an
%% evaluation  of  $x$  and  hence  a  new  map  ${\cal  M}=[\pi  \mapsto
%%   \{x.\sigma\}] $  needs to  be created.  The  ${\cal M}$-map  for the
%% expression $(\SIF~\pi:x~e_1~e_2)$  is a pointwise-union  of the ${\cal
%%   M}$-maps of $e_1$ and $e_2$.   In addition, since the condition also
%% triggers an evaluation of  $x$, the map $[\pi \mapsto \{x.\epsilon\}]$
%% is created and added to  the union.  Also notice that the consequences
%% of  laziness---the entire  expression including  the condition  is not
%% evaluated if  the demand  on it is  $\emptyset$.  This results  in the
%% empty map denoted by $\emptyset_{\cal{M}}$.}
